Exploring clinical and non-clinical(food) Listeria monocytogenes isolates 

### 2025-03-24 code/analysis-code/statistical-analysis.qmd
more statistical analysis

---
title: 
author: 
date: 
output: html_document
---

@@

>>>>

---
title: 
author: 
date: 
output: html_document
---
# loading in packages
```{r}
library(tidyverse)
library(here) 
library(dplyr)
library(skimr)
library(ggplot2)
library(tidymodels)
```
#Load in the data
```{r}
#Path to data. 
data_location <- here::here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

@@

#To do this I have to convert my date variable into a numeric variable.

```{r}
# Load necessary package
library(lubridate)

"# Sample data
mydata <- data.frame(
  date `Create date` = as.Date,
  outcome = clinical  # Yes = 1, No = 0
)"

# Recode Yes = 1, No = 0
mydata$clinical_numeric <- ifelse(mydata$clinical == "Yes", 1, 0)

# Convert date to numeric (days since first observation)
mydata$days_since_start <- as.numeric(difftime(mydata$`Create date`, min(mydata$`Create date`), units = "days"))

# Logistic regression
model <- glm(clinical_numeric ~ days_since_start, data = mydata, family = binomial)

# Model summary
summary(model)

```
```{r}

ggplot(mydata, aes(x = days_since_start, y = clinical_numeric)) +
  geom_jitter(height = 0.05, width = 5, alpha = 0.5) +  
  geom_line(data = new_data, aes(x = days_since_start, y = predicted_prob), color = "blue", size = 1) +
  labs(title = "Logistic Regression: Probability of Clinical Isolates Over Time",
       x = "Days Since Start",
       y = "Probability of Outcome (Yes = 1)") +
  theme_minimal()

```
```{r}
ggplot(mydata, aes(x = days_since_start, fill = factor(clinical_numeric))) +  # Fill by outcome (0 or 1)
  geom_histogram(position = "identity", 
                 alpha = 0.5, 
                 bins = 30, 
                 color = "black") +  # Standard histogram
  labs(title = "Histogram of Outcomes Over Time",
       x = "Days Since Start",
       y = "Count of Outcome (0 or 1)",
       fill = "Outcome") +
  theme_minimal()

```

### 2025-03-24 code/processing-code/processingfile-v1.qmd
more statistical analysis

---
title: "An example cleaning script"
author: "Andreas Handel"
date: "2024-02-07"
output: html_document
---

# Processing script

This contains the same code and comments/information as `processingcode.R`.

This just shows it as Quarto file, to give you an idea how to do it in a setup that combines code and text in a single file.

See the other Quarto file for my currently preferred approach of pulling code from the R script into the Quarto file.

@@

```{r}
library(readxl) #for loading Excel files
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
```

@@

```{r}
lmono2 <- rawdata %>% 
  select(-c(Serovar, Assembly, BioSample))
```
Next I want to reformat the dates so that it removes the timestamp. I will just be keeping the year, month and day
```{r}
lmono2$`Create date` <- as.Date(lmono2$`Create date`)
print(lmono2$`Create date`)
```
Next I will see the range of dates and select a 15 year time frame.
```{r}
lmono3 <- lmono2 %>%
  filter(`Create date` >= as.Date("2010-01-01") & `Create date` <= as.Date("2024-12-31"))
```
Finally I will check for any missing entries, if there are none I am good to move on.
```{r}
any(is.na(lmono3))
lmono3[!complete.cases(lmono3), ]  # Shows me what rows have missing data

>>>>

---
title: " cleaning script"
author: "Alexis Gonzalez"
date: "2024-02-07"
output: html_document
---

@@

```{r}
library(readxl) 
library(dplyr) 
library(tidyr)
library(skimr) 
library(here) 
```

@@

```{r}
lmono2 <- rawdata %>% 
  select(-c(Serovar, Assembly, BioSample))
```
Next I want to reformat the dates so that it removes the timestamp. I will just be keeping the year, month and day
```{r}
lmono2$`Create date` <- as.Date(lmono2$`Create date`)
print(lmono2$`Create date`)
```
Next I will see the range of dates and select a 15 year time frame.
```{r}
lmono3 <- lmono2 %>%
  filter(`Create date` >= as.Date("2010-01-01") & `Create date` <= as.Date("2024-12-31"))
```
Finally I will check for any missing entries
```{r}
any(is.na(lmono3))
lmono3[!complete.cases(lmono3), ]  # Shows me what rows have missing data

### 2025-03-24 products/manuscript/Manuscript.qmd
more statistical analysis

@@

*Provide enough background on your topic that others can understand the why and how of your analysis*

@@

```{}
```

@@

I will briefly use the NNDSS data because unfortunately, it is only availble from 2016 to 2022. The NCBI data containes clinical isolates that date farther back to we will consider those isolates to be "cases". Those cases are what we will compare to the source isolates (environmental, food, etc)

>>>>

Will complete at the end

@@

*Listeria monocytogenes* is a ubiquitous, gram positive, intra-cellular pathogen renowned for its adaptability and resilience (cite). This bacterium can be found in a wide range of environments such as soil, water, animals, and vegetation. Due to its ability to proliferate in food products and low temperatures, it is considered a major food-borne illness. The ingestion of Listeria-infected food may lead to listeriosis, a gastrointestinal illness that causes vomiting, diarrhea, and even death. It is of even greater concern in pregnant individuals, as its forceful invasion tactics allow the bacterium to cross the placental barrier, infecting the fetus; this can lead to miscarriages and stillbirths\[cite\]. Listeriosis causes 1,600 infections and 260 deaths each year with an over 95% hospitalization rate. Listeria can be found in a wide range of meats, produce, and dairy, such as deli meats, soft cheese, and melons. The Centers for Disease Control and Prevention has deemed *Listeria monocytogenes* as a nationally notifiable disease which is a disease that federal agencies, healthcare providers, and laboratories are required to report. All positive isolate data found in the environment, food samples, clinical samples and animal samples are uploaded to the National Center for Biotechnology Information's Pathogen Detection system.

@@

```         
```

@@

I will briefly use the NNDSS data because unfortunately, it is only available from 2016 to 2022. The NCBI data contains clinical isolates that date farther back to we will consider those isolates to be "cases". Those cases are what we will compare to the source isolates (environmental, food, etc)
